Improving the reliability of cognitive task measures: \break A narrative review

### 2022-10-26 manuscript/main.tex
checkpoint progress

% set document class
\documentclass[a4paper,notitlepage,12pt]{article}

@@

% specify the title
\title{Improving the reliability of cognitive task measures: A narrative review}

@@

@@

Cognitive tasks have already demonstrated, and continue to hold, tremendous promise for biological psychiatry. Variability in performance on cognitive tasks provides experimenters with crucial insights into the relationship between individual differences in cognitive processing and behavioral phenomena outside of the laboratory or clinic (e.g. reward sensitivity in a monetary-incentivized task and anhedonia in depression). Elsewhere in psychology, cognitive tasks have been useful in predicting important outcomes such as academic achievement \cite{spiegel2021relations} and cognitive decline \cite{hartshorne2015does}. Cognitive tasks, then, are invaluable tools for building and refining our understanding of psychiatric symptoms and syndromes. For a cognitive task to be useful in this regard, however, it must possess sufficient measurement properties.

@@

We define a cognitive task as any experimental paradigm that measures behavior in order to make inferences about one or more cognitive processes (e.g. Stroop task, delay discounting task, reversal learning task). Cognitive-task measures for characterizing individual differences are descriptive or model-based indicators of behavioral performance such as proportion correct, average response time, and learning rate. The psychometric quality of such a measure can be summarized by three key properties: \emph{discriminating power}, \emph{validity} and \emph{reliability} \cite{kline2015handbook}. The discriminatory power of a task measure describes its ability to measure variability in participants' performance. This is a necessary property of tasks to be used for studying individual differences; where there is no variation in performance, there are no individual differences to study. The validity of a task measure concerns whether it actually measures what it purports to measure. Finally, the reliability of a task measure characterizes the degree to which it consistently rank-orders participants across measurement instances. Put another way, a task measure is reliable if, assuming participants have not changed, it produces the same score for each participant over time. The focus of this review is task reliability, which is a prerequisite for validity.

% P3: definition and Importance of reliability
\subsubsection*{The formal definition of reliability}

In classical test theory, the variance in observed scores on a task measure $X$ is the sum of the true score variance $T$, reflecting real individual differences in the latent construct of interest, and measurement error $E$, i.e., $X = T + E$. The reliability of a measure is defined as the proportion of variance attributable to the true score variance relative to total variance: $\rho = \frac{T}{T + E}$. Thus, reliability quantifies the magnitude of individual differences relative to the noisiness of a task measure; the larger the reliability of a task measure, the more it reflects true individual differences rather than reflecting noise. Reliability is therefore a prerequisite for validity: an unreliable task measure reflects measurement error and not the construct of interest. If this were not reason enough to care about reliability, the observed correlation between two measures (e.g. task performance and self-reported symptom score) is bounded by their individual reliabilities \cite{Spearman1904-mo}: 
\begin{equation}
    \rho_{xy} = \rho_{tt} \sqrt{\rho_{xx'} \cdot \rho_{yy'}}
\end{equation}
where $\rho_{xx'}$ and $\rho_{yy'}$ are the reliabilities of two measures, $x$ and $y$; $\rho_{tt}$ is their true latent correlation; and  $\rho_{xy}$ is their observed correlation. As all reliabilities are $<1$, the reliability of a measure places an upper bound on the maximum observable correlation between itself and a second measure (Figure \ref{fig:fig01}). As an important corollary, as measure reliability decreases, the number of participants required to reliably detect a correlation between two measures increases \cite{Parsons2019-jw}. Thus, poor reliability hampers our ability to investigate associations between cognitive processes, as measured by task performance, and other variables of interest.

\begin{figure}[tp]
    \centering
    \includegraphics[width=1.0\textwidth]{figures/fig01.png}
    \caption{\small The relationship between measure reliability, observed correlations, and statistical power. (A) The maximum expected observed correlation between two measures as a function of their true (latent) correlation and reliability. As the reliability of two measures decreases, so too does their observed correlation. (B) Required sample size for 80\% power to detect true correlations between two measures given their reliability. As the reliability of two measures decreases, the number of participants required to detect an association increases. Even with a large sample size of $N=500$, a measure with reliability $\rho_{tt} = 0.5$ will only allow detection of correlations above $0.3$, which are uncharacteristically high in the field.}
    \label{fig:fig01}
\end{figure}

@@

To further complicate matters, the reliability of a task measure is not absolute -- it reflects interactions between task design, the participants, and the context in which the task is administered. Indeed, task reliability can vary as a function of parameters of the experiment such as the specific stimulus set, number of trials, time limits \cite{paap2016role, cooper2017role}); sample populations (healthy adults, children, psychiatric patients \cite{arnon2020current, cooper2017role}); testing locations (in clinic, online); response modality (desktop, smartphone, virtual reality \cite{pronk2022can, bruder2021reliability}); and scoring or estimation method \cite{Rouder2019-am, haines2020learning}. For example, a cognitive task originally designed for use with an adult population may prove too difficult for children. Their task performance may drop to chance level, thereby minimizing between-participant variance and, as a consequence, task reliability. As a second example, participants completing an experiment online from their homes may experience more distraction than if they participated in the lab. This may increase the measurement error, leading to a concomitant decrease in reliability. Experimenters therefore cannot assume the reliability of a task measure is constant. Ideally researchers would investigate and report the reliability of task measures as part of any analysis or, at the very least, after having made changes to a task or scoring procedure or when administering the task to new sample populations or in new testing contexts.

@@

Although verifying the reliability of cognitive task measures is paramount to individual-differences research, the reliability of task measures is seldom reported \cite{Green2016-xw, Parsons2019-jw}. When they are reported, task measures frequently exhibit lower reliability than what is conventionally considered acceptable for individual-differences research ($\rho_{tt}$ on the order of 0.7 - 0.8) and the reliability regularly achieved by self-report measures. Indeed, many studies have now found that task measures exhibit moderate-to-low reliability \cite{Hedge2018-lf, Frey2017-uz, Enkavi2019-oh, Von_Bastian2020-tm, Nitsch2022-pe, verdejo2021unified}.

@@

As a final point, traditional sum or mean score estimates of performance (e.g., proportion correct responses, mean response time) may substantially underestimate task reliability in cases where some response variability across trials can be accounted for rather than treated as noise \cite{Rouder2019-am, haines2020learning}. These summary scores are noisy (and thus less reliable) estimates of task performance as they neglect trial-by-trial variability. Instead, it is often preferable to use generative or statistical models that account for trial-level variability in behavior. By directly modeling trial-by-trial variability, estimates of participants' performance are effectively de-noised. It is also advisable to use hierarchical models that partially pool data across participants, which in turn further increases the precision of estimates of participants' performance. Together, these methods can substantially boost estimates of reliability \cite{Rouder2019-am, haines2020learning}, as multiple studies have demonstrated  \cite{snijder2022psychometric, sullivan2022enhancing, brown2020improving, waltmann2022sufficient}. For a detailed discussion of generative models in the context of task reliability, see Haines (this issue).

@@

Difference scores deserve special treatment in the context of reliability. Difference scores subtract a measure of a participant's performance in one condition from their performance in another. An example is the classic Stroop interference effect: the average reaction-time difference between congruent and incongruent trials. Difference scores are commonly used because they allow experimenters to isolate particular cognitive processes (e.g. processing cognitive conflict) while controlling for other sources of variance (e.g. perceptual processing, motor ability) through the subtraction of conditions that share that variance. The challenge is that the reliability of a difference score is a function of the reliability of each of its components and the correlation between the components:
\begin{equation}
    \rho_{dd'} = \frac{\sigma^2_x \rho_{xx'} + \sigma^2_y \rho_{yy'} - 2 \rho_{xy} \sigma_x \sigma_y}{\sigma^2_x + \sigma^2_y - 2 \rho_{xy} \sigma_x \sigma_y}
\end{equation}
where $\sigma^2_x$ and $\sigma^2_y$ are the variances of task measures \textit{x} and \textit{y}, $\rho_{xx'}$ and $\rho_{yy'}$ are the reliabilities of task measures \textit{x} and \textit{y}, and $\rho_{xy}$ is the correlation between task measures \textit{x} and \textit{y} \cite{chiou1996reliability}. When the variances of the two measures are equal, this reduces to: 
\begin{equation}
    \rho_{dd'} = \frac{\rho_{xx'} + \rho_{yy'} - 2 \rho_{xy}}{2 - 2 \rho_{xy}}
\end{equation}
From the above equation, one can see that the reliability of a difference score measure is diminished to the extent that its components are correlated. Two measures derived from the same task will often be correlated due to shared domain-general cognitive processes. Thus, it will often (if not always) be the case that difference scores derived from task measures will be less reliable than the average reliability of their components.

@@

The equations above suggest three steps experimenters can take to improve the reliability of difference scores: (1) improve the reliability of the component measures, (2) increase the difference in the magnitude of the variances of the two task measures, and (3) minimize the correlation between the task measures. The first strategy has been our focus so far. The second approach deserves further comment. As discussed elsewhere \cite{chiou1996reliability}, the reliability of a difference score measure increases as the difference between (or the ratio of) the variances of the component measures increases. The intuitive explanation for why this is the case is that, as the variance of one (but not the other) component grows, so to does the proportion of unshared reliable variance. Figure \ref{fig:fig03} shows that even when the correlation of two measures is large, it is possible to achieve acceptable reliability insofar as the variance ratio (i.e. $\sigma_x / \sigma_y$) is high. This speaks to another advantage of increasing the between-participants variability of a measure (insofar that there is an increase in one, but not the other, measure). This fact may help explain the improved reliability of the difference score measures studied by Kucina and colleagues \cite{kucina2022solution}.

\begin{figure}[tp]
    \centering
    \includegraphics[width=1\textwidth]{figures/fig02.png}
    \caption{Difference score reliability as a function of the variances and reliabilities of its component measures. (A) When the variances of the component measures are equal ($\sigma_x / \sigma_y = 1$), larger correlations between the measures substantially diminish the reliability of a difference score measure. (B) When the variances of the component measures are unequal (e.g., here, $\sigma_x / \sigma_y = 4$), even large correlations between the measures impact the reliability of a difference score measure only minimally. Here, the reliabilities of the component measures are assumed to be equal.}
    \label{fig:fig03}
\end{figure}

@@

\section{Discussion}

% outline
% 1) summarize
% 2) validity: three examples, (1) cognitive control + convergent validity, (2) interpretability of task parameters, (3) predictive validity of tasks (impulsivity), (4) ecological validity. maybe even start with rouder paper: some tasks may not be able to made reliable! other challenges of scaling up, repeated testing, naturalistic contexts (cite germine)
% 3) conclusion: pessimism not warranted. requirements: measuring reliability, sharing case studies for improvement, sharing task code (jspsych-demos)

% inspiration 
% The findings presented here convey the importance of scrutinizing the psychometric properties of cognitive tasks, especially those intended for use in individual differences studies. Reliability is not always reported or even examined by researchers, and the present results also demonstrate why variability and reliability should be considered and reported in the context of one another. Future research efforts related to individual differences in cognition could benefit from incorporating psychometric techniques. This may hopefully lead to improvements in the design and evaluation of cognitive tasks for the purposes of individual differences research.

>>>>

% set document class
\documentclass[a4paper,12pt]{article}

@@

% specify the title
\title{Improving the reliability of cognitive task measures: \break A narrative review}

@@

\abstract{Cognitive tasks are capable of providing researchers with crucial insights into the relationship between cognitive processing and psychiatric phenomena. However, many recent studies have found that task measures exhibit poor reliability, hampering their utility for individual differences research. The purpose of this article is to provide a narrative review of approaches to improve the reliability of cognitive task measures. First, we review methods of calculating reliability and discuss some nuances that are specific to cognitive tasks. Then, we introduce a taxonomy of approaches for improving task reliability. Where appropriate, we highlight studies that are exemplary for improving the reliability of specific task measures. We hope that this article can serve as a helpful guide for experimenters who wish to design a new task, or improve an existing task, to achieve sufficient reliability for use in individual-differences research.}

@@

Cognitive tasks have already demonstrated, and continue to hold, tremendous promise for biological psychiatry. Individual differences in performance on cognitive tasks provide researchers with crucial insights into the relationship between cognitive processing and behavioral phenomena in the clinic and in our everyday lives (e.g. reward sensitivity in a monetary-incentivized task and anhedonia in depression). Elsewhere in psychology, cognitive tasks have been useful in predicting important outcomes such as academic achievement \cite{spiegel2021relations} and cognitive decline \cite{hartshorne2015does}. Cognitive tasks, then, are invaluable tools for building and refining our understanding of psychiatric symptoms and syndromes. For a cognitive task to be useful in this regard, however, it must possess sufficient measurement properties.

@@

We define a cognitive task as any experimental paradigm that measures behavior in order to make inferences about one or more cognitive processes (e.g. Stroop task, delay discounting task, reversal learning task). Cognitive-task measures for characterizing individual differences are descriptive or model-based indicators of behavioral performance such as proportion correct, average response time, and learning rate. The psychometric quality of such a measure can be summarized by three key properties: \emph{discriminating power}, \emph{validity} and \emph{reliability} \cite{kline2015handbook}. The discriminatory power of a task measure describes its ability to measure variability in participants' performance. This is a necessary property of tasks to be used for studying individual differences; where there is no variation in performance, there are no individual differences to study. The validity of a task measure concerns whether it actually measures what it intends to measure. Finally, the reliability of a task measure characterizes the degree to which it consistently rank-orders participants across measurement instances. Put another way, a task measure is reliable if, assuming participants have not changed, it produces the same score for each participant over time. The focus of this review is task reliability. 
% P3: definition and Importance of reliability
\subsubsection*{The formal definition of reliability}

In classical test theory, the variance in observed scores on a task measure $x$ is the sum of the true score variance $\sigma^2_T$, reflecting real individual differences in the latent construct of interest, and measurement error $\sigma^2_E$, i.e., $\sigma^2_x = \sigma^2_T + \sigma^2_E$. The reliability of a measure is defined as the proportion of variance attributable to the true score variance relative to total variance: $\rho_{xx'} = \frac{\sigma^2_T}{\sigma^2_T + \sigma^2_E}$. Thus, reliability quantifies the magnitude of individual differences relative to the noisiness of a task measure; the larger the reliability of a task measure, the more it reflects true individual differences rather than reflecting noise. Reliability is therefore a prerequisite for validity: an unreliable task measure reflects measurement error and not the construct of interest. If this were not reason enough to care about reliability, the observed correlation between two measures (e.g. task performance and self-reported symptom score) is bounded by their individual reliabilities \cite{Spearman1904-mo}: 
\begin{equation}
    \rho_{xy} = \rho_{tt} \sqrt{\rho_{xx'} \cdot \rho_{yy'}}
\end{equation}
where $\rho_{xx'}$ and $\rho_{yy'}$ are the reliabilities of two measures, $x$ and $y$; $\rho_{tt}$ is their true latent correlation; and  $\rho_{xy}$ is their observed correlation. As all reliabilities are $<1$, the reliability of a measure places an upper bound on the maximum observable correlation between itself and a second measure (Figure \ref{fig:fig01}). As an important corollary, as measure reliability decreases, the number of participants required to reliably detect a correlation between two measures increases \cite{Parsons2019-jw}. Thus, poor reliability hampers our ability to investigate associations between cognitive processes, as measured by task performance, and other variables of interest.

\begin{figure}[tp]
    \centering
    \includegraphics[width=1.0\textwidth]{figures/fig01.png}
    \caption{\small The relationship between measure reliability, observed correlations, and statistical power. (A) The maximum expected observed correlation between two measures as a function of their true (latent) correlation and reliability. As the reliability of two measures decreases, so too does their observed correlation. (B) Required sample size for 80\% power to detect true correlations between two measures given their reliability. As the reliability of two measures decreases, the number of participants required to detect an association increases. Even with a large sample size of $N=500$, two measures with moderate reliability ($\rho_{xx'} = \rho_{yy'} = 0.5$) will only reliably detect true correlations above $0.3$, which are ostensibly high for the field.}
    \label{fig:fig01}
\end{figure}

@@

To further complicate matters, the reliability of a task measure is not absolute -- it reflects interactions between task design, the participants, and the context in which the task is administered. Indeed, task reliability can vary as a function of parameters of the experiment such as the specific stimulus set, number of trials, time limits \cite{paap2016role, cooper2017role}; sample populations (healthy adults, children, psychiatric patients \cite{arnon2020current, cooper2017role}); testing locations (in clinic, online); response modality (desktop, smartphone, virtual reality \cite{pronk2022can, bruder2021reliability}); and scoring or estimation method \cite{Rouder2019-am, haines2020learning}. For example, a cognitive task originally designed for use with an adult population may prove too difficult for children. Their task performance may drop to chance level, thereby minimizing between-participant variance and, as a consequence, task reliability. As a second example, participants completing an experiment online from their homes may experience more distraction than if they participated in the lab. This may increase measurement error, leading to a concomitant decrease in reliability. Experimenters therefore cannot assume the reliability of a task measure is constant. Ideally researchers would investigate and report the reliability of task measures as part of any individual differences research or, at the very least, after having made changes to a task or scoring procedure or when administering the task to new sample populations or in new testing contexts.

@@

Although verifying the reliability of cognitive task measures is paramount to individual differences research, the reliability of task measures is seldom reported \cite{Green2016-xw, Parsons2019-jw}. When they are reported, task measures frequently exhibit lower reliability than what is conventionally considered the minimum acceptable level for individual-differences research ($\rho_{tt}$ on the order of 0.7 -- 0.8) and the reliability regularly achieved by self-report measures. Indeed, many studies have now found that task measures exhibit moderate-to-low reliability \cite{Hedge2018-lf, Frey2017-uz, Enkavi2019-oh, Von_Bastian2020-tm, Nitsch2022-pe, verdejo2021unified}.

@@

As a final point, traditional sum or mean score estimates of performance (e.g., proportion correct responses, mean response time) may substantially underestimate task reliability \cite{Rouder2019-am, haines2020learning}. These summary scores are contaminated by trial-level noise that, in the absence of sufficiently large number of trials, increase the measurement error in (and thus diminish the reliability of) these measures of task performance. Instead, it is often preferable to use generative or statistical models that account for trial-level variability in behavior. By directly modeling trial-by-trial variability, estimates of participants' performance are effectively de-noised. It is also advisable to use hierarchical models that partially pool data across participants, which in turn further increases the precision of estimates of participants' performance. Together, these methods can substantially boost estimates of reliability \cite{Rouder2019-am, haines2020learning}, as multiple studies have demonstrated  \cite{snijder2022psychometric, sullivan2022enhancing, brown2020improving, waltmann2022sufficient}. For a detailed discussion of generative models in the context of task reliability, see Haines (this issue).

@@

Difference scores deserve special treatment in the context of reliability. Difference scores subtract a measure of a participant's performance in one condition from their performance in another. An example is the classic Stroop interference effect: the average reaction-time difference between congruent and incongruent trials. Difference scores are commonly used because they allow experimenters to isolate particular cognitive processes (e.g. processing cognitive conflict) while controlling for other sources of variance (e.g. perceptual processing, motor ability) through the subtraction of conditions that share that variance. The challenge is that the reliability of a difference score is a function of the reliability of each of its components \textit{and} the correlation between the components:
\begin{equation}
    \rho_{dd'} = \frac{\sigma^2_x \rho_{xx'} + \sigma^2_y \rho_{yy'} - 2 \rho_{xy} \sigma_x \sigma_y}{\sigma^2_x + \sigma^2_y - 2 \rho_{xy} \sigma_x \sigma_y}
\end{equation}
where $\sigma^2_x$ and $\sigma^2_y$ are the variances of task measures \textit{x} and \textit{y}, $\rho_{xx'}$ and $\rho_{yy'}$ are the reliabilities of task measures \textit{x} and \textit{y}, and $\rho_{xy}$ is the correlation between task measures \textit{x} and \textit{y} \cite{chiou1996reliability}. When the variances of the two measures are equal, this reduces to: 
\begin{equation}
    \rho_{dd'} = \frac{\rho_{xx'} + \rho_{yy'} - 2 \rho_{xy}}{2 - 2 \rho_{xy}}
\end{equation}
From the above equation, one can see that the reliability of a difference score measure is diminished to the extent that its components are correlated. Two measures derived from the same task will often be correlated due to shared domain-general cognitive processes. Thus, it will often (if not always) be the case that difference scores derived from task measures will be less reliable than the average reliability of their components.

@@

The equations above suggest three steps experimenters can take to improve the reliability of difference scores: (1) improve the reliability of the component measures, (2) increase the relative difference between the variances of two task measures, and (3) minimize the correlation between the task measures. The first strategy has been our focus so far. The second approach deserves further comment. As discussed elsewhere \cite{chiou1996reliability}, the reliability of a difference score measure increases as the difference between (or the ratio of) the variances of the component measures increases. Intuitively this is because, as the variance of one (but not the other) component grows, so too does the proportion of unshared reliable variance. Figure \ref{fig:fig03}A shows that even when the correlation of two measures is large, it is possible to achieve acceptable reliability insofar as the ratio of the variances (i.e. $\sigma_x / \sigma_y$) is sufficiently greater or less than 1. (Also worth noting is that component variances essentially function as weights as in a weighted average, such that difference score reliability more reflects the reliability of the component measure with the larger variance; Figure \ref{fig:fig03}B.) This speaks to another advantage of increasing the between-participants variability of a measure of performance in one experiment condition insofar that there is not an accompanying increase in perormance variability in a second condition. This may explain why making more difficult the incongruent trials in the tasks designed by Kucina and colleagues improved the reliability of their difference score measures \cite{kucina2022solution}.

\begin{figure}[tp]
    \centering
    \includegraphics[width=1\textwidth]{figures/fig02.png}
    \caption{\small Difference score reliability as a function of the variances and reliabilities of its component measures. (A) Difference score reliabilities as the ratio of the component measures' variances ($\sigma_x / \sigma_y$) increase, but the component reliabilities are held fixed ($\rho_{xx'} = \rho_{yy'} = 0.8$). When the variances of the component measures are equal ($\sigma_x / \sigma_y = 1$), larger correlations between the measures substantially diminish the reliability of a difference score measure. When the variances are unequal ($\sigma_x / \sigma_y > 1$), even large correlations between the measures are less deleterious for reliability. (B) Difference score reliabilities as the component reliabilities change, but the ratio of the component measures' variances is fixed ($\sigma_x / \sigma_y = 2$). When the variances are unequal, the reliability of a difference score reflects more the component measure with the larger variance.}
    \label{fig:fig03}
\end{figure}

@@

\section{Conclusion}

In this brief article, we have reviewed issues and research regarding cognitive task-measure reliability. Specifically, we introduced a taxonomy of approaches for improving cognitive task-measure reliability, highlighting studies that are exemplary for improving the reliability of specific task measures. We hope we have made clear the importance of calculating (and reporting) the reliability of task measures intended for use in psychiatric research. We also hope we have provided a useful guide for experimenters who wish to design a new task, or improve an existing task, to study individual differences in cognitive processing.

Before concluding, we must emphasize that reliability is but one of many important considerations in the design and evaluation of cognitive task measures. Task measures may be reliable but show poor convergent validity \cite{snijder2022psychometric, eckstein2022interpretation}, raising questions about whether we actually measuring the constructs we intended to measure. Similarly, task measures may be reliable but exhibit poor ecological validity \cite{Steiner2021-oq}, thus being poor proxies for cognition in real-world settings. Task measures may also be reliable but show poor predictive validity \cite{verdejo2021unified}, explaining little (unique) variance in other variables we care about (e.g. symptoms, treatment response). Finally, there are many other practical considerations (e.g. duration, engagement, accessibility) to ensure cognitive tasks are able to be deployed successfully to the clinic or naturalistic contexts at scale \cite{germine2021toward}.

Despite the challenges of making task-measures reliable and valid, we remain optimistic about their current and future use in biological psychiatry. We firmly believe that, with further research efforts towards developing, documenting, and sharing reliable task paradigms, our field can make great strides towards understanding, predicting, and ultimately relieving psychiatric illness.

\section{Code availability}

The code used to generate the figures in this manuscript are publicly available at \break \url{https://github.com/nivlab/biopsych-reliability-review}.

\section{Acknowledgments}

This research was funded in part by a National Science Foundation Graduate Research Fellowship.

\section{Disclosures}

The authors have no conflicts of interest to declare.
